Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2018-02-24 paper.tex
conformation selection rewritten

\begin{abstract}
Chemical interactions between proteins and other molecules (ligands) take place in specific locations, called active sites, that are often deeply buried inside the protein structure.
These active sites are accessible through one or more void paths, called tunnels.
Nowadays, tunnels are mostly computed using Voronoi diagrams which approximate the ligand by a bounding sphere and 
the traversability of the tunnels is determined based only on the radial bottlenecks.
This can lead to highly imprecise results.
To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in the high-dimensional configuration space.
We propose a modification of the Rapidly Exploring Random Tree planner for the purpose of trajectory computations for flexible ligands.
To cope with the narrow passage problem, the random samples are generated around a tunnel detected by a Voronoi-based approach.
%which serves us for better localization of the searched space.
The flexibility of the ligand is modeled using a fixed set of known conformations. 
%which does not increase the number of dimensions of the configuration space.
The proposed method has been evaluated in a virtual screening scenario to estimate the traversability of tunnels for several ligands.
\end{abstract}

@@

The widely used approaches to evaluate the ligand traversability rely on the tunnel detection.
Most of the available tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3,sehnal2013mole}.
Tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and the list of surrounding residues, whose physico-chemical properties are crucial for assessing of the feasibility of the protein-ligand interactions.
The tunnels are detected either on a single conformation of the protein, or, on a set of protein conformations, generated by MD simulations.
Computationally less demanding approaches for generating protein conformations are based on constrained geometric simulations, such as
ROCK~\cite{lei2004sampling}, FRODA~\cite{wells2005constrained}, and tCONCOORD~\cite{seeliger2007geometry}.
Examples of utilization of protein tunnels can be found in the recently published state-of-the-art report about the analysis and visualization of biomolecular cavities~\cite{Krone_2016}.

@@

The RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in the difficult regions, as the growth of the configuration tree towards the random samples may be blocked by obstacles. %~\cite{vonasekphd}.
To prevent this blocking, the DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for the expansion to a small ball. 
The radius of this ball is set to infinity for new nodes, and decreases to a predefined radius if the node cannot be successfully expanded.
An automatic adaptation of this radius was proposed in the ADD-RRT~\cite{jailletADRRT}.
%The disadvantage of DD-RRT is the sensitivity to the predefined radius.
%Therefore, the authors of ADD-RRT~\cite{jailletADRRT} proposed to adapt the ball radius according to the success rate of the expansion.
To attract the tree towards a given region, random samples have to be generated there only if the tree can expand to this region.
In~\cite{kardossRRTKK}, the probability of sampling is increased in several waypoints close to narrow passages, but without specifying how to find these waypoints.
The generalization of~\cite{kardossRRTKK} is the guided sampling, where the whole path is used to generate samples in the configuration space.
The path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on the solution of a relaxed version of the problem~\cite{bayazitIRC}.

@@

%Another techniques to sample the narrow passages dense enough is to retract the random samples towards regions that are believed
%to be located in narrow passages~\cite{zhangRetraction,lee2012srrrt}.
%the random sample $\qrand$ is connected to the tree if the line segment  from $\qnear$ to $\qrand$ is collision-free. Otherwise, the retraction-step is performed.
%The task of the retraction step is to find a contact configuration around $\qnear$ that minimizes the distance to $\qrand$.
In the Retraction-based RRT~\cite{zhangRetraction}, the tree is retracted along the boundary of the obstacles in the configuration space.
This requires the analysis of the contact space, which leads to the generalized penetration depth computation of high complexity~\cite{he2016efficient}. %\cite{he2016efficient} in RAL!! + zhang2008fast
The Selective Retraction-based RRT~\cite{lee2012srrrt} avoids the expensive growth in the open areas of the configuration space and focuses the retractions only on the narrow passages.
The Obstacle-based RRT~\cite{amatoOBRRT} utilizes several expansion procedures based on, e.g., random vectors, obstacle vectors, or medial axis.
In MARRT~\cite{denny2014marrt}, newly generated samples are pushed towards the medial axis.
%The operator can also select a suitable planner for the region.<F12>
%Therefore, many algorithms uses a heuristic to compute the samples near boundary of the configuration space~\cite{amatoOBPRM,amatoOBRRT}.

@@

\section{Proposed Method}
We propose a modification of RRT to compute trajectories for non-spherical flexible ligands through protein tunnels.
The input of the proposed method is a set of tunnels computed using Voronoi diagrams, and the task is to find trajectories for a ligand along these tunnels leading from the protein surface to the inner buried active site.
The tunnels are detected on selected protein conformations, resulting, e.g., from MD simulations or from a tool providing the conformation sampling
~\cite{lei2004sampling,wells2005constrained,seeliger2007geometry}.
%The tunnels are detected in a sequence of time frames of MD simulations.
%capturing protein dynamic behavior. 
%The frames are computed using MD simulations.

@@

Due to the protein dynamics, the tunnels move and merge, and their bottleneck can change in each time step.
It can be expected that the tunnels in MD simulations without ligands are rather narrow.
However, the tunnels can adapt their shape to the passing ligands and vice versa.
%Therefore, it can be expected that the detected tunnels are rather narrow.
In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can potentially serve as the transportation path for the ligand.
Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates the shrinking of the ligand size.
In our solution, we simulate this by shrinking the atom radii of the ligand, used also by~\cite{cortes2010simulating,guieysse2008structure}.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere representing an atom is given by its van der Waals radius.
Protein tunnel is described by a sequence of collision-free spheres 
$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
$c_i \in \R^3$ their 3D position, and $r_i > 0$ the maximum collision-free radius of a sphere centered at $c_i$. 
Tunnels serving as an input for the proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.
%Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
%or conformations may be generated on-the-fly in the presence of the receptor binding cavity,[14] 
%or with full rotational flexibility of every dihedral angle using fragment based docking.[15] 
%Force field energy evaluation are most often used to select energetically reasonable conformations,[16] but 
%knowledge-based methods have also been used.[17]
%13 = https://link.springer.com/article/10.1007/BF00123666
%The proteins are dense structures and the tunnels are typically narrow.
%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
The ligand is modeled using the set $\L$ of its conformations.
The candidate conformations are prepared by discrete sampling of degrees of freedom of the ligand.
Each candidate conformation is ranked according to its energy and only low-energy conformations are used in $\L$.
%The conformations are prepared using the Rosetta tool and ranked according their potential energy. Only conformations
%with energy lower than a user-defined threshold are used.
%The flexibility of the ligand is modeled using the set $\L$ of its conformations.
%We assume that the library of all possible conformations in a considered resolution is available
%(e.g., ~\cite{dunbrack}) 
%and we later show how to select a subset of conformations suitable for the motion planning.
To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
In our solution, we use the discrete set of scales, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale, and $\sdelta$ is the minimal difference between two scales.

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$ is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., the geometric center or a selected atom), rotation around $x$, $y$, and $z$ axes,
index of the conformation $l\in \L$, and scale $s \in \S$.
All possible configurations form the configuration space $\C$ and the collision-free configurations
form the subspace $\CF \subseteq \C$.
A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with protein atoms.
We assume that the ligand can change from conformation $l_1$ in configuration $q_1$ to conformation $l_2$ in configuration $q_2$
if both $q_1$ and $q_2$ are collision-free and if the distance from $q_1$ to $q_2$ is smaller than the resolution of the planner ($\varepsilon=0.05$~\AA).

@@

The second modification is the expansion procedure (Alg.~\ref{alg::expand}). % which expands the tree from $\qnear$ towards $\qrand$.
%The expansion towards $\qrand$ is crucial for the growth of the tree towards the unexplored areas of the configuration space.
The traditional straight-line expansion (i.e., new configuration is constructed on the line from $\qnear$ to $\qrand$ in the distance
$\varepsilon$ from $\qnear$) more likely leads to collisions, especially in the narrow void space.
We expand the tree considering all $\L$ conformations.
%The proposed expansion attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with the highest scale.
First, the maximal scale $\smax \in \S$ is considered and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
and tested for collisions.
The random samples are generated similarly as in case of $\qrand$ samples, the translation part $(x,y,z)$ is generated around $\qnear$.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
By testing first the largest scale ($\smax$), the expansion procedure prefers expanding the tree by larger scales, and the
smaller scales are used only in difficult (narrow) regions.

@@

Each conformation from $\L$ is tested during the tree expansion. %in order to approach $\qrand$.
By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
Using a large number of conformations also increases the time complexity of the expansion step.
Selection of a suitable set of conformations $\L$ is therefore crucial.
Depending on the ligand, some conformations can be considered as more packed, i.e., the atoms in such conformations are closer to each other, but 
the atoms can also form a more elongated shape (Tab.~\ref{fig::m040c}).
From the motion planning point of view, both types of conformations are useful.
The longer conformations can better fit in the narrow passage, while the packed conformations can easily rotate on the spot.
%In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
%packed conformations can be used.
%\vspace{-3mm}

The pool of available conformations $\L$ consists of all conformations (up to some resolution) with the energy lower than a user-specified threshold, which can still contain too many conformations (thousands or even more).
We further select a subset by computing how the atoms of the ligand are spread around its geometric center.
The distance of centers of ligand atoms to the geometric center of the ligand is computed, and the deviation of these
distances then characterizes the spread.
Low deviation is characteristic for the packed conformations, while larger deviations correspond to the elongated conformations.
To select $n$ conformations from all $|\L|$ available conformations, they are sorted according to the deviation.
Each $k-$th conformation from the sorted list, where $k=\lfloor{|\L| / n}\rfloor$, is used.

@@

The methods \RA\ and \RB\ differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configuration $\qnew$ that approaches $\qrand$.
The retraction step is realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.05$~\AA\ using 20 samples.
A collision-free sample that maximally approaches $\qrand$ is selected and added to the tree, and the retraction continues from this sample.
The retraction step is repeated 5 times.
The \RB\ method can therefore expand the tree by at most 5 new nodes per configuration, while \RA\ extends the tree only by one configuration.
The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for the nearest-neighbor search and the time to solve these queries is significantly smaller than the computing collision detection.
The parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is the same.
For each conformation and tested scale, the \RA\ method attempts to find $m=100$ collision-free configurations, and \RB\ attempts to perform $5 \times 20$ collision-detection queries.
Other parameters of the \RB\ planner were the same as for \RA.
The planning was terminated after $\Imax$ iterations or if the tree approached the active site to the distance less than 2~\AA\ (i.e., the 
geometric center of the ligand was closer than 2~\AA\ from the active site).

@@

%The case was the analysis of Leukotriene A4 hydrolase (PDB ID 4L2L).
The task of the second analysis was to find the trajectory for the ligand 4-(4-benzylphenyl)-1,3-thiazol-2-amine into the active
site of A4 hydrolase (PDB ID 4L2L).
%This ligand is a known inhibitor of the protein 4L2L.
%Based on the location of inhibitor inside the protein we assume that the inhibitor is able 
The tCONCOORD tool~\cite{seeliger2007geometry} was used to generate 1000 conformations of the protein 4L2L (without the inhibitor).
The initial PDB structure was minimized by GROMACS 4.0.7~\cite{hess2008gromacs} using the input parameters recommended on the example page for tCONCOORD. 
The protein conformations were generated using the default settings for tCONCOORD in the PERT mode. % which only perturbs the starting configuration of atoms instead of complete randomization.
The tunnels leading to the active site defined by residues 134 (Alanine) and 311 (Phenylalanine) were computed by CAVER 3.0, using the probe 0.9~\AA.
All 1000 conformations contained at least 5 tunnels, more than 900 conformations contained 9 tunnels.
The detected the tunnels slightly differ due to the conformational changes.
Therefore, the ligand traversability towards the active site was computed for the first three tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA\, and 1.19~\AA\ with the average lengths 22.3~\AA, 25.3~\AA\, and 28.4~\AA, respectively).
The conformations of the ligand were prepared in the same way as in the previous section.
From the available $|\L|=736$ conformations, 100 were selected using the `S' strategy.

>>>>

\begin{abstract}
Chemical interactions between proteins and other molecules (ligands) take place in specific locations, called active sites, that are often deeply buried inside the protein structure.
These active sites are accessible through one or more void paths, called tunnels.
Nowadays, tunnels are mostly computed using Voronoi diagrams which approximate the ligand by a bounding sphere and 
the traversability of the tunnels is determined based only on the radial bottlenecks.
This can lead to imprecise results as the shape of the ligand is not considered.
To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in the high-dimensional configuration space.
We propose a modification of the Rapidly Exploring Random Tree planner for the purpose of trajectory computations for flexible ligands.
To cope with the narrow passage problem, the random samples are generated around a tunnel detected by a Voronoi-based approach.
%which serves us for better localization of the searched space.
The flexibility of the ligand is modeled using a fixed set of known conformations. 
%which does not increase the number of dimensions of the configuration space.
The proposed method has been evaluated in a virtual screening scenario to estimate the traversability of tunnels for several ligands.
\end{abstract}

@@

The widely used approaches to evaluate the ligand traversability rely on the tunnel detection.
Most of the available tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3,sehnal2013mole}.
Tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and the list of surrounding residues, whose physico-chemical properties are crucial for assessing of the feasibility of the protein-ligand interactions.
The tunnels are detected either on a single conformation of the protein, or, on a set of protein conformations, generated by MD simulations.
Computationally less demanding approaches for generating possible protein conformations are based on constrained geometric simulations, such as
ROCK~\cite{lei2004sampling}, FRODA~\cite{wells2005constrained}, and tCONCOORD~\cite{seeliger2007geometry}.
Examples of utilization of protein tunnels can be found in the recently published state-of-the-art report about the analysis and visualization of biomolecular cavities~\cite{Krone_2016}.

@@

The RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in the difficult regions, as the growth of the configuration tree towards the random samples may be blocked by obstacles. %~\cite{vonasekphd}.
To prevent this blocking, the DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for the expansion to a small ball. 
The radius of this ball is set to infinity for new nodes, and decreases to a predefined radius if the node cannot be successfully expanded.
An automatic adaptation of this radius was proposed in the ADD-RRT~\cite{jailletADRRT}.
%The disadvantage of DD-RRT is the sensitivity to the predefined radius.
%Therefore, the authors of ADD-RRT~\cite{jailletADRRT} proposed to adapt the ball radius according to the success rate of the expansion.
To attract the tree towards a given region, random samples have to be generated there only if the tree can expand to this region.
In~\cite{kardossRRTKK}, the probability of sampling is increased in several waypoints close to narrow passages, but without specifying how to find these waypoints.
The generalization of~\cite{kardossRRTKK} is the guided sampling, where the whole path is used to generate samples in the configuration space.
The guiding path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on the solution of a relaxed version of the problem~\cite{bayazitIRC}.

@@

%Another techniques to sample the narrow passages dense enough is to retract the random samples towards regions that are believed
%to be located in narrow passages~\cite{zhangRetraction,lee2012srrrt}.
%the random sample $\qrand$ is connected to the tree if the line segment  from $\qnear$ to $\qrand$ is collision-free. Otherwise, the retraction-step is performed.
%The task of the retraction step is to find a contact configuration around $\qnear$ that minimizes the distance to $\qrand$.
In the Retraction-based RRT~\cite{zhangRetraction}, the tree is retracted along the boundary of the obstacles in the configuration space.
%This requires the analysis of the contact space, which leads to the generalized penetration depth computation of high complexity~\cite{he2016efficient}. %\cite{he2016efficient} in RAL!! + zhang2008fast
The Selective Retraction-based RRT~\cite{lee2012srrrt} avoids the expensive growth in the open areas of the configuration space and focuses the retractions only on the narrow passages.
The Obstacle-based RRT~\cite{amatoOBRRT} utilizes several expansion procedures based on, e.g., random vectors, obstacle vectors, or medial axis.
In MARRT~\cite{denny2014marrt}, newly generated samples are pushed towards the medial axis.
%The operator can also select a suitable planner for the region.<F12>
%Therefore, many algorithms uses a heuristic to compute the samples near boundary of the configuration space~\cite{amatoOBPRM,amatoOBRRT}.

@@

\section{Proposed Method}
We propose a modification of RRT to compute trajectories for non-spherical flexible ligands through protein tunnels.
The input of the proposed method is a set of tunnels computed using Voronoi diagrams, and the task is to find trajectories for a ligand along these tunnels leading from the protein surface to the inner buried active site.
The tunnels are detected on selected protein conformations, 
    resulting, e.g., from MD simulations or from a tool providing the protein conformation sampling
~\cite{lei2004sampling,wells2005constrained,seeliger2007geometry}.
%The tunnels are detected in a sequence of time frames of MD simulations.
%capturing protein dynamic behavior. 
%The frames are computed using MD simulations.

@@

Due to the protein dynamics, the tunnels move and merge, and their bottleneck can change in each time step.
It can be expected that the tunnels in MD simulations without ligands are rather narrow.
However, the tunnels can adapt their shape to the passing ligands and vice versa.
%Therefore, it can be expected that the detected tunnels are rather narrow.
In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can potentially serve as the transportation path for the ligand.
Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates the shrinking of the ligand size.
In our solution, we simulate this by shrinking the atom radii of the ligand, which is used also by~\cite{cortes2010simulating,guieysse2008structure}.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere representing an atom is given by its van der Waals radius.
Protein tunnel is described by a sequence of collision-free spheres 
$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
$c_i \in \R^3$ their 3D position, and $r_i > 0$ the maximum collision-free radius of a sphere centered at $c_i$. 
Tunnels serving as an input for the proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.
%Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
%or conformations may be generated on-the-fly in the presence of the receptor binding cavity,[14] 
%or with full rotational flexibility of every dihedral angle using fragment based docking.[15] 
%Force field energy evaluation are most often used to select energetically reasonable conformations,[16] but 
%knowledge-based methods have also been used.[17]
%13 = https://link.springer.com/article/10.1007/BF00123666
%The proteins are dense structures and the tunnels are typically narrow.
%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
The ligand is modeled using the set $\L$ of its conformations.
The conformations are prepared by discrete sampling of degrees of freedom of the ligand. % (e.g., using the Rosetta3 tool).
Each candidate conformation is ranked according to its energy and only low-energy conformations are used in $\L$.
%The conformations are prepared using the Rosetta tool and ranked according their potential energy. Only conformations
%with energy lower than a user-defined threshold are used.
%The flexibility of the ligand is modeled using the set $\L$ of its conformations.
%We assume that the library of all possible conformations in a considered resolution is available
%(e.g., ~\cite{dunbrack}) 
%and we later show how to select a subset of conformations suitable for the motion planning.

To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
In our solution, we use the discrete set of scales, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale, and $\sdelta$ is the minimal difference between two scales.

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$ is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., the geometric center or a selected atom), rotation around $x$, $y$, and $z$ axes,
index of the conformation $l\in \L$, and scale $s \in \S$.
All possible configurations form the configuration space $\C$ and the collision-free configurations
form the subspace $\CF \subseteq \C$.
A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with protein atoms.
We assume that the ligand can change from conformation $l_1$ in configuration $q_1$ to conformation $l_2$ in configuration $q_2$
if both $q_1$ and $q_2$ are collision-free and if the distance from $q_1$ to $q_2$ is smaller than the resolution $\varepsilon$ of the planner ($\varepsilon=0.05$~\AA).

@@

The second modification is the expansion procedure (Alg.~\ref{alg::expand}). % which expands the tree from $\qnear$ towards $\qrand$.
%The expansion towards $\qrand$ is crucial for the growth of the tree towards the unexplored areas of the configuration space.
%The traditional straight-line expansion (i.e., new configuration is constructed on the line from $\qnear$ to $\qrand$ in the distance
%$\varepsilon$ from $\qnear$) more likely leads to collisions, especially in the narrow void space.
We expand the tree considering all $\L$ conformations.
%The proposed expansion attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with the highest scale.
First, the maximal scale $\smax \in \S$ is considered and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
and tested for collisions.
The random samples are generated similarly as in case of $\qrand$ samples, the translation part $(x,y,z)$ is generated around $\qnear$.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
By testing first the largest scale ($\smax$), the expansion procedure prefers expanding the tree by larger scales, and the
smaller scales are used only in difficult (narrow) regions.

@@

Each conformation from $\L$ is tested during the tree expansion. %in order to approach $\qrand$.
By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
Using a large number of conformations also increases the time complexity of the expansion step.
The pool of available conformations $\L$ consists of all conformations (up to some resolution) with the energy lower than a user-specified threshold, which can still contain too many conformations (thousands or even more).

In the case of large sets $\L$, only a subset of the available ligand conformations can be used in the planning.
To select the subset, the conformations are ranked using the spread of the atoms around the geometric center of the ligand.
The distance of centers of ligand atoms to the geometric center of the ligand is computed, and the deviation of these
distances then characterizes the spread.
Low deviation is characteristic for `packed' conformations, while larger deviations correspond to the elongated conformations (Tab.~\ref{fig::m040c}).
To select the subset of $n$ representative conformations from all $|\L|$ available conformations, they are sorted according to the deviation.
Each $k-$th conformation from the sorted list, where $k=\lfloor{|\L| / n}\rfloor$, is used.
The selected conformations are then used in the planning process.

%Selection of a suitable set of conformations $\L$ is therefore crucial.
%Depending on the ligand, some conformations can be considered as more packed, i.e., the atoms in such conformations are closer to each other, but%the atoms can also form a more elongated shape (Tab.~\ref{fig::m040c}).
%From the motion planning point of view, both types of conformations are useful.
%The longer conformations can better fit in the narrow passage, while the packed conformations can easily rotate on the spot.
%In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
%packed conformations can be used.
%\vspace{-3mm}
%We further select a subset by computing how the atoms of the ligand are spread around its geometric center.

@@

The methods \RA\ and \RB\ differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configuration $\qnew$ that approaches $\qrand$.
The retraction step is realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.05$~\AA\ using 20 samples.
A collision-free sample that maximally approaches $\qrand$ is selected and added to the tree, and the retraction continues from this sample.
The retraction step is repeated 5 times.
The \RB\ method can therefore expand the tree by at most 5 new nodes per ligand conformation, while \RA\ extends the tree only by one new node.
The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for the nearest-neighbor search and the time to solve these queries is significantly smaller than the computing collision detection.
The parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is the same.
For each conformation and tested scale, the \RA\ method attempts to find $m=100$ collision-free configurations, and \RB\ attempts to perform $5 \times 20$ collision-detection queries.
Other parameters of the \RB\ planner were the same as for \RA.
The planning was terminated after $\Imax$ iterations or if the tree approached the active site to the distance less than 2~\AA\ (i.e., the 
geometric center of the ligand was closer than 2~\AA\ from the active site).

@@

%The case was the analysis of Leukotriene A4 hydrolase (PDB ID 4L2L).
The task of the second analysis was to find the trajectory for the ligand 4-(4-benzylphenyl)-1,3-thiazol-2-amine into the active
site of A4 hydrolase (PDB ID 4L2L, 9666 atoms).
%This ligand is a known inhibitor of the protein 4L2L.
%Based on the location of inhibitor inside the protein we assume that the inhibitor is able 
The tCONCOORD tool~\cite{seeliger2007geometry} was used to generate 1000 conformations of the protein 4L2L (without the inhibitor).
The initial PDB structure was minimized by GROMACS 4.0.7~\cite{hess2008gromacs} using the input parameters recommended on the example page for tCONCOORD. 
The protein conformations were generated using the default settings for tCONCOORD in the PERT mode. % which only perturbs the starting configuration of atoms instead of complete randomization.
The tunnels leading to the active site defined by residues 134 (Alanine) and 311 (Phenylalanine) were computed by CAVER 3.0, using the probe 0.9~\AA.
All 1000 conformations contained at least 5 tunnels, more than 900 conformations contained 9 tunnels.
The ordering of the detected tunnels slightly differ in different conformations.
Therefore, the ligand traversability towards the active site was computed for the first three widest tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA\, and 1.19~\AA\ with the average lengths 22.3~\AA, 25.3~\AA\, and 28.4~\AA, respectively).
The conformations of the ligand were prepared in the same way as in the previous section.
From the available $|\L|=736$ conformations, 100 were selected using the `S' strategy.
